---
title: "filling_package_skeleton_with_life"
author: "Katja Danielzik"
date: "2024-08-07"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# raw data 
will be worked on later

# data
import already pre-processed intracellular metabolite concentrations with one
column specifying the experimental condition (in our case radiation dose) and
one column specifying time (numerical)

## intracellular concentrations
```{r}
intra <- read_csv("/media/home/5-1-package pipeline/data/intra_corrected_cell_counts.csv")
# save as rda file
use_data(intra)
# write documentation
promptData(intra)
```
## KEGG-database 
these files are summaries of the pull request from KEGG via KEGGREST
```{r}
# total background information
modules_compounds  <- readr::read_csv("/media/home/5-1-package pipeline/data/modules_compounds.csv")
devtools::use_data(modules_compounds)
devtools::promptData(modules_compounds)

# experimental metabolites mapped to modules
metabolite_modules <- readr::read_csv("/media/home/5-1-package pipeline/data/metabolite_modules.csv")
devtools::use_data(metabolite_modules)
devtools::promptData(metabolite_modules)

# suggest KEGGREST as package for developers to test the code or obtain own datasets
devtools::use_package("KEGGREST","suggests",min_version=TRUE)
devtools::document()
```

That is all we need for now.

## Compiled Stan models
Pre-compiled Stan models make it much easier and faster for the user
```{r}
load("/media/home/5-1-package pipeline/compiled_dynamics_model.RData")
use_data(compiled_dynamics_model)

```



