---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# MetaboDynamics

<!-- badges: start -->
<!-- badges: end -->

MetaboDynamics provides a framework of Bayesian models for robust and easy
analysis of longitudinal Metabolomics Data. It takes concentration tables as input an provides robust estimation of mean concentrations, functional enrichment
analysis employing the KEGG database and comparison of dynamic clusters of 
different biological conditions.

## Installation

You can install the development version of MetaboDynamics from [GitHub](https://github.com/) with:

``` r
# install.packages("pak")
pak::pak("KatjaDanielzik/MetaboDynamics")
```

## Example

This is a basic example which shows you how to solve a common problem:

```{r example}
library(MetaboDynamics)
## basic example code
```


